Response to reviewers

### 2025-01-06 Manuscript/ecography_reponse_letter_2.tex
up reviewer 1

\textcolor{blue}{Line 89-90: I think that some references are necessary here.}

@@

@@

@@

@@

@@

\textcolor{blue}{Line 339: I would recommend placing figure 2 after the first results paragraph.}

\textcolor{blue}{Figure 3: I would recommend placing figure 3 in the results section. }

@@

@@

@@

@@

\textcolor{blue}{Lines 128-135: The capacity of the algorithm to simulate the stressor strongly depends on the distribution of the data and the assumption of independence in the simulated dimensions (as should be the case in the simulation example).
Given this, it is unclear why a single trait is simulated as multidimensional.
This approach feels unnecessarily complex, potentially confounding the reader and introducing covariation that should be tested.
To clarify, a trait is typically defined as a single measurable characteristic of an organism, particularly when considering the definition referenced in Violle et al. (2007).
The concept of multidimensionality in trait spaces usually applies when multiple traits are combined into a matrix, with each column representing a distinct trait or a principal component summarizing trait variation.
However, it is not common practice to describe a single trait using multiple variables (i.e., columns or dimensions).

Regarding the statement in Lines 132–135:
\textit{“These aspects are often expressed in one dimension but can in fact be described in any number of dimensions. For example, for trait defined as “leaf insertion angle”, this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).”}
While this explanation clarifies the authors’ definition of a “multidimensional trait,” it appears to complicate the methodology.
If traits are often expressed in a single dimension (e.g., angle in degrees), I do not see the necessity to simulate them as “a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.” (lines 128-129 If I understood the text correctly, this approach results in a trait space built around a single trait defined by multiple dimensions.

Instead, it would be more intuitive and broadly applicable to simulate 2, 4, or 8 separate traits.
These traits, together, would define a 2-, 4-, or 8-dimensional trait space.
Since these traits would be uncorrelated, they would correspond directly to the same number of principal components, aligning with the definition of dimensions provided in Supplementary Table 1.
Moreover, simulating a single multidimensional trait may inadvertently introduce covariation across dimensions.
For example, using “leaf insertion angle” as an illustration, the trigonometric relationships of the triangle’s sides and angles inherently interconnect mathematically (e.g., through sine, cosine, and tangent).
This interconnection could introduce correlation in the variance, even under Brownian assumptions.
While this correlation may not be inherently incorrect or distinct from real ecological data, the assumption of independence in the simulation should be explicitly tested.
Correlated dimensions might create clustering in the trait space, potentially biasing the application of stressors in the simulation context. 
Finally, simulating independent traits using Brownian motion without fixing the variance could also address the issue of homogeneity of variance across dimensions (Lines 463–464).
Homogeneity of variance may introduce similar bias in the simulated mechanisms and it is far from approximating a set of independent trait or dimensions.
I belive that simulating independent trait may mitigate that concern.}

>>>>

\textcolor{blue}{Line 89-90: I think that some references are necessary here.}
We changed this sentence and merged it with the next one after discovering some previously missed literature:

\textit{Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g. \cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines @@@.

@@

We rephrased the sentence to:

\textit{In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e. observations adjacent to each other in the trait space are more likely compared to common ones).} Lines @@@

@@

We replaced the dash by a comma.

@@

We removed the brackets.

@@

We've now explained why we chose p=3:

\textit{here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D} Lines @@@.

@@

We fixed the type.

\textcolor{blue}{Line 339: I would recommend placing figure 2 after the first results paragraph.}
\textcolor{blue}{Figure 3: I would recommend placing figure 3 in the results section. }

The placing of figures is done automatically here by the compiler.
But we'll make sure to forward this suggestion to the copy editors after the manuscript is accepted.

@@

We remove the word ``say''.

@@

We've now simplified the sentence to:

\textit{Our results show that the loss of  functional diversity for several metrics was not more than expected by chance.
This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).} Lines @@@

@@

We've now specified the following:

\textit{For example, we can follow the suggestions from \cite{mammola2021concepts} highlighting which kind of methodological pipeline best supports which type of metric of interest (Table 3).} Lines @@@

@@

\textcolor{blue}{Lines 128-135: The capacity of the algorithm to simulate the stressor strongly depends on the distribution of the data and the assumption of independence in the simulated dimensions (as should be the case in the simulation example).
Given this, it is unclear why a single trait is simulated as multidimensional.
This approach feels unnecessarily complex, potentially confounding the reader and introducing covariation that should be tested.
To clarify, a trait is typically defined as a single measurable characteristic of an organism, particularly when considering the definition referenced in Violle et al. (2007).
The concept of multidimensionality in trait spaces usually applies when multiple traits are combined into a matrix, with each column representing a distinct trait or a principal component summarizing trait variation.
However, it is not common practice to describe a single trait using multiple variables (i.e., columns or dimensions).}

\textcolor{blue}{Regarding the statement in Lines 132–135:
\textit{“These aspects are often expressed in one dimension but can in fact be described in any number of dimensions. For example, for trait defined as “leaf insertion angle”, this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).”}
While this explanation clarifies the authors’ definition of a “multidimensional trait,” it appears to complicate the methodology.
If traits are often expressed in a single dimension (e.g., angle in degrees), I do not see the necessity to simulate them as “a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.” (lines 128-129 If I understood the text correctly, this approach results in a trait space built around a single trait defined by multiple dimensions.}

\textcolor{blue}{Instead, it would be more intuitive and broadly applicable to simulate 2, 4, or 8 separate traits.
These traits, together, would define a 2-, 4-, or 8-dimensional trait space.
Since these traits would be uncorrelated, they would correspond directly to the same number of principal components, aligning with the definition of dimensions provided in Supplementary Table 1.
Moreover, simulating a single multidimensional trait may inadvertently introduce covariation across dimensions.
For example, using “leaf insertion angle” as an illustration, the trigonometric relationships of the triangle’s sides and angles inherently interconnect mathematically (e.g., through sine, cosine, and tangent).
This interconnection could introduce correlation in the variance, even under Brownian assumptions.
While this correlation may not be inherently incorrect or distinct from real ecological data, the assumption of independence in the simulation should be explicitly tested.
Correlated dimensions might create clustering in the trait space, potentially biasing the application of stressors in the simulation context. 
Finally, simulating independent traits using Brownian motion without fixing the variance could also address the issue of homogeneity of variance across dimensions (Lines 463–464).
Homogeneity of variance may introduce similar bias in the simulated mechanisms and it is far from approximating a set of independent trait or dimensions.
I belive that simulating independent trait may mitigate that concern.}

### 2025-01-06 Manuscript/main_ecography_R2.tex
R2 INIT
up reviewer 1

While the first part of this framework (``what?'') has been studied extensively in the last decade, this was often undertaken outside of the ``what, how and why'' framework.
Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home to both populations and another one home to only one of the population (the process - ``how?'').
In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations because it might not give information on whether the two populations actually interact in some way - a very likely condition for competition to take place.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.

@@

\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.
For example $D_{20}$ is the distance that excludes any $n_{20}$ pairs of points that are a distance of at least $D_{20}$ from each other, $D_{40}$ is the distance that excludes $n_{40}$ pairs, etc. (with $D_{20} > D_{40} > D_{60} > D_{80}$).
Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
We chose this algorithm to approximate the facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
I.e. pairs of observations that share similar trait combinations are more likely to remain (i.e. observations adjacent to each other in the trait space).

\item Shifting space: by removing the required amount of data from a distance (radius) $\rho'$ of the observation with the maximum numerical value on all dimensions.
This is similar to the decreasing size algorithm but instead of choosing $\rho$ to be the radius from the centre of the trait space, $\rho'$ is a radius from the ``top right corner'' of the trait space.
In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).
This approximates our \textbf{filtering mechanism}.
We chose this algorithm to approximate a flitering mechanism assuming a stressor that could increase the probability of extinction for observations further from some trait combination optimum.
This is similar to our equalising mechanism approximation but with the optimum not being in the center of the trait space (a region with a high density of observations) but with the optimum being in a corner of the trait space (a region with low density).

\item Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
This is done by selecting $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0()} function in \texttt{R};][]{silverman1986density}.
Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$. 
Where $p$ is a factor increasing the scaling power of the algorithm (here we used $p=3$).
In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.
This could approximate the fact that observations that have similar trait combinations are more likely to go extinct due to competition (e.g. using the same resources). 
\end{itemize}

@@

The full trait dataset comprised $118$ bird species, all native species known to have existed on these islands over the last $125,000$ years.
$55$ species are known to have gone extinct prior to $1500$ CE and $26$ after $1500$ CE.
Note that we used the extinct species data as presented in \cite{matthews2023global}, and thus we do not include de Hawaiian species declared extinct by the IUCN in $2024$.
For our analyses, we focused on two time periods: (i) the avifauna present in $1500$ CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-$1500$ CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).

@@

For example, say one is interested in looking at changes in functional diversity using a divergence metric calculated with the probability density method (e.g. \texttt{TPD::REND}).
They can apply this metric to their trait space and get two relative metric scores of, say, $0.7$ and $0.65$ (pattern) for two levels of extinction (process).
The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.
In our simulations, if the mechanism approximated was the equalising one, the pattern can clearly be interpreted as a change in functional diversity, but if the mechanism is a facilitation one, the pattern can be meaningless (i.e. not clearly distinguishable from a null mechanism - Fig. \ref{Fig:simulation_results} shows this example by contrasting row 5, columns 1 and 2).
In this made up example, the procedure is done in reverse order where the the pattern is specified first and then the process and the mechanism (``What process and mechanism are needed to interpret pattern X?'').
In Box 2 we suggest an approach that specifies the process and the mechanism first (``how do I interpret pattern X under a specific process and mechanism''; see Box 2).

@@

Our results indicating that the loss of functional diversity (based on several metrics) was not more than expected by chance when focusing on all extinctions could be due to the fact that although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines (approximatively $58$\%), the majority of global known anthropogenic extinctions (approx. $75$\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
However, in Hawaii, where the same proportion of known species were passerines (approx. $58$\%), the opposite is true: the majority of extinctions (approx. $60$\%) have been of passerine species.
It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
Given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines, our results indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e. relative to random extinctions).
In addition, it is worth stressing that these analyses included marine species (which constitute less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.

@@

Once the one has clearly defined the pattern they want to capture (\textbf{what}), we believe it gets easier to pick up the type of metric that would be the most adequate.
For example, we can follow the suggestions from \cite{mammola2021concepts} (Table \ref{Tab:box2}).

@@

%TG: TODO: check the pre-1500 story from email from Tom and Maria

>>>>

Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g. \cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home to both populations and another one home to only one of the population (the process - ``how?'').
In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations because it might not give information on whether the two populations actually interact in some way - a very likely condition for competition to take place.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.

@@

\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.
For example $D_{20}$ is the distance that excludes any $n_{20}$ pairs of points that are a distance of at least $D_{20}$ from each other, $D_{40}$ is the distance that excludes $n_{40}$ pairs, etc. (with $D_{20} > D_{40} > D_{60} > D_{80}$).
Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
We chose this algorithm to approximate the facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e. observations adjacent to each other in the trait space are more likely compared to common ones).

\item Shifting space: by removing the required amount of data from a distance (radius) $\rho'$ of the observation with the maximum numerical value on all dimensions.
This is similar to the decreasing size algorithm but instead of choosing $\rho$ to be the radius from the centre of the trait space, $\rho'$ is a radius from the ``top right corner'' of the trait space.
In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions, in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).
This approximates our \textbf{filtering mechanism}.
We chose this algorithm to approximate a flitering mechanism assuming a stressor that could increase the probability of extinction for observations further from some trait combination optimum.
This is similar to our equalising mechanism approximation but with the optimum not being in the center of the trait space (a region with a high density of observations) but with the optimum being in a corner of the trait space (a region with low density).

\item Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
This is done by selecting $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0} function in \texttt{R};][]{silverman1986density}.
Each discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$. 
Where $p$ is a factor increasing the scaling power of the algorithm (here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D).
In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
This approximates our \textbf{competition mechanism}, where observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.
This could approximate the fact that observations that have similar trait combinations are more likely to go extinct due to competition (e.g. using the same resources). 
\end{itemize}

@@

The full trait dataset comprised $118$ bird species, all native species known to have existed on these islands over the last $125,000$ years.
$55$ species are known to have gone extinct prior to $1500$ CE and $26$ after $1500$ CE.
Note that we used the extinct species data as presented in \cite{matthews2023global}, and thus we do not include Hawaiian species declared extinct by the IUCN in $2024$.
For our analyses, we focused on two time periods: (i) the avifauna present in $1500$ CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-$1500$ CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).

@@

For example, say one is interested in looking at changes in functional diversity using a divergence metric calculated with the probability density method (e.g. \texttt{TPD::REND}).
They can apply this metric to their trait space and get two relative metric scores of $0.7$ and $0.65$ (pattern) for two levels of extinction (process).
The interpretation of this score as a change in functional diversity will depend on the worker's assumption of which mechanism affected the trait space.
In our simulations, if the mechanism approximated was the equalising one, the pattern can clearly be interpreted as a change in functional diversity, but if the mechanism is a facilitation one, the pattern can be meaningless (i.e. not clearly distinguishable from a null mechanism - Fig. \ref{Fig:simulation_results} shows this example by contrasting row 5, columns 1 and 2).
In this made up example, the procedure is done in reverse order where the the pattern is specified first and then the process and the mechanism (``What process and mechanism are needed to interpret pattern X?'').
In Box 2 we suggest an approach that specifies the process and the mechanism first (``how do I interpret pattern X under a specific process and mechanism''; see Box 2).

@@

Our results show that the loss of  functional diversity for several metrics was not more than expected by chance.
This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
However, in Hawaii, where the same proportion of known species were passerines (approx. $58$\%), the opposite is true: the majority of extinctions (approx. $60$\%) have been of passerine species.
It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
Given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines, our results indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e. relative to random extinctions).
In addition, it is worth stressing that these analyses included marine species (which constitute less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.

@@

Once the one has clearly defined the pattern they want to capture (\textbf{what}), we believe it gets easier to pick up the type of metric that would be the most adequate.
For example, we can follow the suggestions from \cite{mammola2021concepts} highlighting which kind of methodological pipeline best supports which type of metric of interest (Table \ref{Tab:box2}).

@@
